Spatial Reasoning and Data Displays

### 2015-06-27 Paper/InfoVis2015Revision/TestingVisualAptitude.Rnw
study procedure, start

% \hhnote{XXX this procedural description needs a few more pieces: order of different tests (pilot study showed that order did not matter), 
% total experimental time, experimental setup: in a lab with paper copies;  
% 
% This information is actually present in the "Experimental procedure" section below. This section is just discussing lineups.}

In this study, we utilize three blocks of statistical lineups; each block consists of 5-7 datasets, each displayed using three or four types of plots (examples of the visualization types presented in each block of lineups are shown in figure \ref{fig:lineups}). For example, in lineup set 1, the same five datasets are displayed as a dot plot, density plot, box plot, and histogram. Each block of lineups are a subset of stimuli used in experiments designed to compare the visual significance of different plot types, however, our purpose in this study is not to determine the efficacy of different types of plots. Rather, our goal is to compare the cognitive demands required to successfully evaluate these lineups, which display variance difference, mean difference, or violations of normality using common statistical plots.

@@

\paragraph{Experimental Procedure}
All tasks (cognitive and lineups) were provided to participants on paper, to conform to the procedure used in~\cite{ekstrom1976manual}. The order of tasks was the same for each participant due to the constraints of in-person testing; internal pilot tests did not indicate an effect of task order.

@@

In addition to these tests, participants were asked to complete a questionnaire which includes questions about colorblindness, mathematical background, self-perceived verbal/mathematical/artistic skills, time spent playing video games, and undergraduate major. These questions are designed to assess different factors which may influence a participant's skill at reading graphs and performing spatial tasks.

After each test or block of lineups, participants were given 1-2 minutes to rest and recover; the order of tests was designed to prevent fatigue on a single task. The duration of testing for each participant was approximately 1 hour. All test documents (with copyrighted material redacted) are available as supplemental materials.

>>>>

In this study, we utilize three blocks of statistical lineups; each of these blocks consists of 5-7 datasets, that are displayed in three or four types of plots (examples of the visualization types presented in each block of lineups are shown in figure \ref{fig:lineups}). For example, in the lineups of set 1, the same five datasets are displayed as a dot plot, density plot, box plot, and histogram. Each block of lineups are a subset of stimuli used in experiments designed to compare the visual significance of different plot types, however, our purpose in this study is not to determine the efficacy of different types of plots. Rather, our goal is to compare the cognitive demands required to successfully evaluate these lineups, which display variance difference, mean difference, or violations of normality using common statistical plots.

@@

\paragraph{Experimental Procedure}
All tasks (cognitive and lineups) were provided to participants on paper \newtext{in a supervised lab setting}, to conform to the procedure used in~\cite{ekstrom1976manual}. The order of tasks was the same for each participant due to the constraints of in-person testing; results from an internal pilot tests with a counterbalanced design did not indicate an effect of task order, nor was there any indication in the results of an interaction/learning effect from one task to the next.

@@

In addition to these tests, participants completed a questionnaire on personal background and demographics, including questions about colorblindness, mathematical training, self-perceived verbal/mathematical/artistic skills, time spent playing video games, and undergraduate major. These questions are designed to assess different factors which may influence a participant's skill at reading graphs and performing spatial tasks.

After each test or block of lineups, participants were given 1-2 minutes to rest and recover; the order of tests was designed to prevent fatigue on a single task \newtext{(which led us to interweave lineup evaluations)}. The duration of testing for each participant was approximately 1 hour. \newtext{None of the participants who showed up for their scheduled session asked to leave early.} A set of test documents (with copyrighted material redacted) is available as supplemental materials.
